Mergeable Types

### 2017-02-27 paper/dali.tex
edits to related work
Edits.
Sec. 2
edits

% Listings Code
%---------------
\newcommand{\lstml}{
\lstset{ %
language=ML, % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={module, @@deriving, not, : , /\\}
}}
\lstnewenvironment{ocaml}
    { % \centering
			\lstml
      \lstset{}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Distributed applications often replicate data to improve
  availability and fault tolerance.  But, the use of replication leads
  to a significantly more complex programming model.  To ensure a
  sensible measure of consistency among different replicas of an
  object, applications are typically structured so that the operations
  applied to a replicated object on one node are subsequently
  communicated to other nodes where they can be re-applied locally.
  As a result, the degree of restructuring and care necessary to
  transform a sequential or shared-memory concurrent application to a
  meaningful (replicated) distributed one can be substantial.

  In this paper, we present a new interpretation of replicated types
  that overcome these concerns.  Our declarative programming model
  called \name\ treats replication as a form of versioning, in which
  consistency is enforced through programmer-supplied \emph{merge}
  functions.  Distributed state is now viewed in terms of a tree of
  immutable object versions, with merge functions used implicitly by
  the underlying runtime system to periodically reconcile local
  versions to yield a consistent global state.  Thus, our model frees
  the programmer from having to think of low-level operational notions
  related to distribution and replication, and therefore does not
  entail any restructuring of application logic.

  We show how any ML datatype can be enriched with mergeable
  semantics, formalize a rich concurrency model based on such types ,
  and describe an instantiation of OCaml equipped with these features.
  Our implementation is integrated with a content-addressable
  distributed storage layer that allows any OCaml program to be easily
  transformed into an efficient, highly available, distributed
  application.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{keyword1, keyword2, keyword3}  %% \keywords is optional

@@

%% Acknowledgments
% \begin{acks}                            %% acks environment is optional
%                                         %% contents suppressed with 'anonymous'
%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
%   %% acknowledge financial support and will be used by metadata
%   %% extraction tools.
%   This material is based upon work supported by the
%   \grantsponsor{GS100000001}{National Science
%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%   conclusions or recommendations expressed in this material are those
%   of the author and do not necessarily reflect the views of the
%   National Science Foundation.
% \end{acks}

>>>>

% Listings Code
%---------------
\newcommand{\lstml}{
\lstset{ %
language=ML, % choose the language of the code
basicstyle=\footnotesize\ttfamily,       % the size of the fonts that are used for the code
keywordstyle=\color{Bittersweet},
% numbers=left,                   % where to put the line-numbers
numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
stepnumber=1,                   % the step between two line-numbers. If it is 1 each line will be numbered
numbersep=5pt,                  % how far the line-numbers are from the code
showspaces=false,               % show spaces adding particular underscores
showstringspaces=false,         % underline spaces within strings
showtabs=false,                 % show tabs within strings adding particular underscores
% frame=single,                   % adds a frame around the code
tabsize=2,                      % sets default tabsize to 2 spaces
captionpos=b,                   % sets the caption-position to bottom
breaklines=true,                % sets automatic line breaking
breakatwhitespace=false,        % sets if automatic breaks should only happen at whitespace
commentstyle=\itshape\color{MidnightBlue},
%escapeinside={\%*}{*)},         % if you want to add a comment within your code
morekeywords={module, when, @@deriving, not, : , /\\}
}}
\lstnewenvironment{ocaml}
    { % \centering
			\lstml
      \lstset{}%
      \csname lst@setfirstlabel\endcsname}
    { %\centering
      \csname lst@savefirstlabel\endcsname}

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  Distributed applications often replicate data to improve
  availability and fault tolerance.  But, the use of replication leads
  to a significantly more complex programming model.  To ensure a
  sensible measure of consistency among different replicas of an
  object, applications are typically structured so that the operations
  applied to a replicated object on one node are subsequently
  communicated to other nodes where they can be re-applied locally.
  Oftentimes, the degree of restructuring and care necessary to
  transform a sequential or shared-memory concurrent application to a
  meaningful (replicated) distributed one can be substantial.

  In this paper, we present a new interpretation of replicated types
  that overcome these concerns.  Our declarative programming model
  called \name\ treats replication as a form of versioning, in which
  consistency is enforced through programmer-supplied \emph{merge}
  functions.  Distributed state is now viewed in terms of a tree of
  immutable object versions, with merge actions used explicitly by the
  application or implicitly by the underlying runtime system to
  periodically reconcile local versions to yield a consistent global
  state.  Thus, our model frees the programmer from having to think of
  low-level operational notions related to distribution and
  replication, and therefore does not entail any restructuring of
  application logic to enable distributed programming.

  We show how any ML datatype can be enriched with mergeable
  semantics, formalize a rich concurrency model based on such types,
  and describe an instantiation of OCaml equipped with these features.
  Our implementation is integrated with a content-addressable
  distributed storage layer that allows any OCaml program to be easily
  transformed into an efficient, highly available, distributed
  application.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008.10011009.10010177</concept_id>
<concept_desc>Software and its engineering~Distributed programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011008.10011024.10011028</concept_id>
<concept_desc>Software and its engineering~Data types and structures</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10011007.10011006.10011039.10011311</concept_id>
<concept_desc>Software and its engineering~Semantics</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10003752.10010124.10010125.10010130</concept_id>
<concept_desc>Theory of computation~Type structures</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~Distributed programming languages}
\ccsdesc[500]{Software and its engineering~Data types and structures}
\ccsdesc[300]{Software and its engineering~Semantics}
\ccsdesc[300]{Theory of computation~Type structures}

%% End of generated code

%% Keywords
%% comma separated list
\keywords{Replicated Datatypes, Eventual Consistency, Availability, CRDTs, Versions, ML}  %% \keywords is optional

@@

\input{related}
%% Acknowledgments
% \begin{acks}                            %% acks environment is optional
%                                         %% contents suppressed with 'anonymous'
%   %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
%   %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
%   %% acknowledge financial support and will be used by metadata
%   %% extraction tools.
%   This material is based upon work supported by the
%   \grantsponsor{GS100000001}{National Science
%     Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
%   No.~\grantnum{GS100000001}{nnnnnnn} and Grant
%   No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
%   conclusions or recommendations expressed in this material are those
%   of the author and do not necessarily reflect the views of the
%   National Science Foundation.
% \end{acks}

### 2017-02-27 paper/intro.tex
edits

Real-world distributed programs are challenging to write and maintain
because they often conflate two distinct mechanisms.  The first
concerns the expression of application logic - how do we define
computations that are robust in the presence of distributed
communication among concurrently executing threads of control?  The
second deals with system concerns - how do we express notions of
visibility, replication, and consistency when the nodes participating
in such computations may be geographically distributed and the
networks that connect them unreliable?  To simplify reasoning in such
complicated environments, programming models often make strong
assumptions on the guarantees provided by an implementation (such as
serializability~\cite{Serializability} or strong
consistency~\cite{CDE+12,DNN+15}) that may inadvertently mask,
restrict, or ignore important albeit unpleasant realities (e.g.,
network partitions~\cite{Brewer2000,Gilbert2002}). Given these
unwarranted assumptions, program behavior is often difficult to
predict and verify.  On the other hand, when these features are
explicitly exposed to the programmer, e.g., by requiring that
applications be written in terms of specialized distributed data
structures~\cite{Burckhardt2014,BFL+12,SPB+11} or control
primitives~\cite{Calm}, simplicity, composability, and
ease-of-reasoning can suffer.

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instanteously witnessed on another, which may in the process of
simultaneously attempting to perform its own \C{Add} or \C{Mult}
action.  While synchronizing the activities of all nodes to ensure at
most one such operation is performed at a time is impractical,
designating a single node to hold the counter state eschewing
replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given its sensitivity to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removig coordination altogether by
simply replicating counter state without having any supporting
consistency protocol is an equally infeasible approach.

A method often adopted~\cite{SPB+11,Burckhardt2014} to address this
tension is to re-define a datatype's operations to return
\emph{effects} instead of values.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all replicas to
incorporate the effects of the original
operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
with operations re-defined to returns effects.  Observe that the
\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
when applied at a replica (see \C{apply} in the figure), adds \C{x} to
the local counter value.  Note, however, that \C{add} and \C{mult} are
not commutative operations - assuming two replicas have the same
initial counter value of 0, applying the effect of adding 3 and then
multiplying 5 on one replica yields 15, while applying the effect of
first multiplying 5 and then adding 3 yields 3 on the other.  Thus,
implementations must be carefully written to take the lack of
commutativity into account when defining how effects are applied at a
replica; in the figure, multiplication is expressed in terms of
addition to avoid the kind of undesirable behavior described above.

@@

counter value, assuming computation on the counter quiesces.  While
transmitting effects provide a low-level operational basis for

@@

replication and the other not, breaks backward compatibility.  Just as
significantly, its non-trivial construction must be developed in
different guises for every distributed data structure used by the
application.

@@

    derived to yield a new merged state of 25.  Observe that there is
    no explicit synchronization or coordination among versions.  In
    the figure, red circles represent counter state produced through
    merging, and blue circles represent state produced by applying
    counter operations on a node-local version.  }

@@

with one branch for each replica.  Thus, each branch represents

@@

state produced by a counter's operations on a node recorded along the
node's local branch.  Now, to generate a globally consistent view of a
counter, we only need to define a merge operation that explains how to
combine two local versions to produce a new version that reflects both
their states.

@@

In this paper, we develop a programming model that realizes a monadic
version control system centered around data, rather than file,
coherence.  The \name monad lets programmers write and compose
concurrent computations around multiple (implicit) versions of a
mergeable datatype, an ordinary ML datatype additionally equipped with
a \emph{merge} function responsible for deriving a consistent global
state from a collection of local versions of that state.  A
computation progresses by forking replicating new versions of existing
versions on geo-distributed nodes, creating new versions along a an
existing \emph{branch}, or by merging branch tips.

Our model allows ML programmers to get the benefits of realizing
highly available (low-latency) distributed computation, while
continuing to enjoy the comforts of high-level reasoning and the
familiarity of using standard libraries.  Notably, while \name's
programming model makes no explicit reference to any specific
operational manifestation of a distributed system (e.g., programmers
do not need to explicitly manage replicas), we demonstrate that it can
be nonetheless efficiently realized on existing geo-replicated
distributed systems.

@@

    \item We describe the \name library that transparently adds
      persistence and replication features to a mergeable type. \name
      comes with a programming model that brings to bear the power and
      flexibility of a version control system to the administration of
      replicated data. \name hides the complexity of version control
      behind a monad, and exposes its functionality via a simple API
      that lets programmers define and compose distributed ML
      computations around such data.  Issues related to replication
      only manifest implicitly in the definition of a merge operation
      that defines coherence among different instances of replicated
      state.

@@

    \item We describe an implementation of \name atop the Irmin
      persistent store~\cite{irmin}, a content-addressable storage
      library for OCaml, along with case studies and experimental
      results that justify its practical utility.
\end{itemize}

>>>>

Real-world distributed programs are challenging to write and maintain
because they often conflate two distinct mechanisms.  The first
concerns the expression of application logic - how do we define
computations that are robust in the presence of distributed
communication among concurrently executing threads of control?  The
second deals with system concerns - how do we express notions of
visibility, replication, and consistency when the nodes participating
in such computations may be geographically distributed and the
networks that connect them unreliable?  To simplify reasoning in such
complicated environments, programming models often make strong
assumptions on the guarantees provided by an implementation (such as
serializability~\cite{Serializability} or strong
consistency~\cite{CDE+12,DNN+15}) that may inadvertently mask,
restrict, or ignore important albeit unpleasant realities (e.g.,
network partitions~\cite{Brewer2000,Gilbert2002}). When unwarranted
assumptions are made, program behavior is often difficult to predict
and verify.  On the other hand, when these features are explicitly
exposed to the programmer, e.g., by requiring that applications be
written in terms of specialized distributed data
structures~\cite{Burckhardt2014,BFL+12,SPB+11} or control
primitives~\cite{Calm}, simplicity, composability, and
ease-of-reasoning can suffer.

To illustrate how these kinds of problems manifest in practice,
consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instanteously witnessed on another, which may be in the process of
simultaneously attempting to perform its own \C{Add} or \C{Mult}
action.  While synchronizing the activities of all nodes to ensure at
most one such operation is performed at a time is impractical,
designating a single node to hold the counter state eschewing
replication altogether (as in a client-server
configuration~\cite{Armstrong}), is also an undesirable solution,
given the sensitivity of such architectures to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removig coordination altogether by simply
replicating counter state without having any supporting consistency
protocol is an equally infeasible approach.

A method often adopted to address these concerns is to re-define a
datatype's operations to return \emph{effects} instead of
values~\cite{SPB+11,Burckhardt2014}.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all replicas to
incorporate the effects of the original
operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
with operations re-defined to returns effects.  Observe that the
\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
when applied at a replica (see \C{apply} in the figure), adds \C{x} to
the local counter value.  Note, however, that \C{add} and \C{mult} are
not commutative operations - assuming two replicas have the same
initial counter value of 0, applying the effect of adding 3 and then
multiplying 5 on one replica yields 15, while applying the effect of
first multiplying 5 and then adding 3 yields 3 on the other.  Such
scenarios are possible in a distributed system because there are no
coherence guarantees on the order in which effects are received by
different nodes.  Thus, implementations must be carefully written to
take the lack of commutativity into account when defining how effects
are applied at a replica; in the figure, multiplication is expressed
in terms of addition to avoid the kind of undesirable behavior
described above.

@@

counter value, assuming updates to the counter eventually quiesces.
While transmitting effects provide a low-level operational basis for

@@

replication and the other not, breaks backward compatibility with the
original state-based implementation.  Just as significantly, its
non-trivial construction must be developed in different guises for
every distributed data structure used by the application.

@@

    derived to yield a new merged state of 25.  There is no other form
    of synchronization or coordination among versions except through
    merging.  In the figure, red circles represent counter state
    produced through merging, and blue circles represent state
    produced by applying counter operations on a node-local version.
  }

@@

with one branch for each replica.  Thus, every branch represents

@@

state produced by the computation performed over a counter on a node
recorded along the node's local branch for the counter.  Now, to
generate a globally consistent view of a counter, we only need to
define a merge operation that explains how to combine two local
versions to produce a new version that reflects both their states.

@@

In this paper, we develop a programming model that realizes a monadic
version control system centered around data, rather than file,
coherence.  The \name monad lets programmers write and compose
concurrent computations around multiple (implicit) versions of a
mergeable datatype, an ordinary ML datatype additionally equipped with
a \emph{merge} function responsible for deriving a consistent global
state from a collection of local versions of that state.  A
computation progresses by forking (i.e., replicating) new versions of
existing versions on geo-distributed nodes, allowing
synchronization-free local computation to proceed on those nodes,
creating new versions along an existing \emph{branch} that represents
updated local instances, and by merging branches to realize global
consistency.

Our model allows ML programmers to get the benefits of achieving
highly available (low-latency) distributed computation, while
continuing to enjoy the comforts of high-level reasoning and the
familiarity of using standard libraries already provided by the
language.  Notably, while \name's programming model makes no explicit
reference to any specific operational manifestation of a distributed
system (e.g., programmers do not need to explicitly manage replicas),
we demonstrate that it can be nonetheless efficiently realized on
existing geo-replicated distributed systems.

@@

    \item We describe \name programming model that brings to bear the
      power and flexibility of a version control system to the
      administration of replicated data. \name hides the complexity of
      version control behind a monad, and exposes its functionality
      via a simple API that lets programmers define and compose
      distributed ML computations around such data.  Issues related to
      replication only manifest implicitly in the definition of a
      merge operation that defines coherence among different instances
      of replicated state.

@@

    \item We describe an implementation of the \name library that
      transparently adds persistence and replication features to a
      mergeable type, and which sits atop the Irmin persistent
      store~\cite{irmin}, a content-addressable storage library for
      OCaml.  We also present case studies and experimental results
      that justify the practical utility of our approach.
\end{itemize}
